---
title: "Ukraine Support Tracker Data Import"
author: "Nathan Ormond"
date: "`r Sys.Date()`"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load Required Libraries

```{r load-libraries}
# Install required packages if missing
if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
if (!requireNamespace("janitor", quietly = TRUE)) install.packages("janitor")

# Load libraries
library(readxl)
library(tidyverse)
library(here)
library(janitor)
```

## 2. Initial Data Exploration

First, let's check what sheets exist in the Excel file:

```{r explore-sheets}
xlsx_path <- here("data", "raw", "f319e1c8-5654-4cd6-b4c7-5722ae437d30-Ukraine_Support_Tracker_Release_21.xlsx")

# List all sheets in the Excel file
sheet_names <- excel_sheets(xlsx_path)
print(sheet_names)
```

## 3. Examining Key Data Sheets

With 91 sheets in the Excel file, we need to be strategic about our exploration. Let's focus on sheets that likely contain the most important data for our analysis.

```{r examine-key-sheets}
# Define key sheets that are likely to contain the main data
key_sheets <- c(
  "Bilateral Assistance, MAIN DATA",  # This appears to be the primary dataset
  "Country Summary (€)",              # Summary in Euros
  "Country Summary ($)",              # Summary in USD
  "Aggregates by Country Group"       # Group-level aggregations
)

# Loop through these key sheets and display their structure
for (sheet in key_sheets) {
  cat("\n\n### Sheet:", sheet, "\n")
  
  # Try to read the first few rows
  tryCatch({
    sample_data <- read_excel(xlsx_path, sheet = sheet, n_max = 5)
    print(glimpse(sample_data))
    
    # Also show column names for easier reference
    cat("\nColumn names:\n")
    print(colnames(sample_data))
    
    # Show dimensions
    cat("\nDimensions (rows x columns):", dim(sample_data)[1], "x", dim(sample_data)[2], "\n")
    
    # Show data types for each column
    cat("\nColumn data types:\n")
    print(sapply(sample_data, class))
  }, error = function(e) {
    cat("Error reading sheet:", e$message, "\n")
  })
}
```


## 4. Creating Ukraine Support Chart

Now let's create a stacked bar chart showing aid contributions by country and type, similar to the reference image.

```{r load-country-summary}
# Load the Country Summary sheet in Euros
country_summary <- read_excel(xlsx_path, sheet = "Country Summary (€)")

# Take a look at the structure of the data
glimpse(country_summary)

# Display column names to identify the right columns for our chart
colnames(country_summary)
```

Based on the column names we see, let's prepare the data for plotting:

```{r}
# First, let's read the sheet without skipping rows to see its full structure
raw_data <- read_excel(xlsx_path, sheet = "Country Summary (€)")

# Look at the first 15 rows to understand the layout
head(raw_data, 15)

# Check for any row that contains country names we recognize
country_rows <- which(raw_data[[1]] %in% c("United States", "Germany", "United Kingdom", "Japan", "France"))
print(paste("Rows with country names:", paste(country_rows, collapse = ", ")))

# Let's see what rows 7-12 look like (adjusting based on what we find above)
raw_data[7:12, 1:7]  # Show first 7 columns


# identify what each column represents
# check the headers by viewing rows 6-7 for all columns
raw_data[6:7, 1:10]  # Show first 10 columns to understand the headers

# specific columns across more rows
raw_data[7:20, c(1, 4, 5, 6, 7)]  # First column plus likely aid columns

# Look at rows 5-7 to see all the column headers
raw_data[5:7, 4:10]  # Focus on columns that likely contain our data of interest

``` 

```{r}
```